---
pagetitle: Review of probability I
output: 
  revealjs::revealjs_presentation:
    incremental: false
    theme: solarized
    self_contained: false
    # reveal_plugins: ["menu","notes","chalkboard"]
    reveal_plugins: ["menu"]
    highlight: pygments
    center: true
    transition: none
    background_transition: none 
    reveal_options:
      # chalkboard:
      #   theme: whiteboard
      #   toggleNotesButton: true
      #   toggleChalkboardButton: true
      menu:
        numbers: true
      slideNumber: true
      previewLinks: false
    fig_caption: true
    css: mystyle.css
    
--- 

<section>

<h1>Random variables and distributions</h1>

Based on Stock and Watson, ch. 2

<br>

<h2>[Jesper Bagger](mailto:jesper.bagger@rhul.ac.uk)</h2>

<h3>EC2203 | Royal Holloway | 2020/21</h3>



</section>

# Outline

- Random variables

- Probability distributions

# Random variables

## Randomness

- A random process is a definite outcome that cannot be predicted with certainty

  - Coin flip
  
  - Student test scores (for example, your EC2203 mark)
  
  - Waiting time in the College shop
  
  - The FTSE100 index
  
  - Salary in your first job
  
- We use **random variables** to analyze random processes
  
  
## Random variables

- Provides a **numerical** summary of a random outcome

- A random variable assigns a **real number** to every possible outcome of a random process

- A random variable takes **different values** depending on the outcomes of the random process under study


# Distributions

## The distribution of a random variable

- The probability that, say, $X=x$ is the **proportion of times $X=x$ in many independent realizations** of $X$

- The **distribution** of a random variable specifies the probability of realizing each value of the random variable

- Distinguish b/w **discrete** and **continuous** random variables


## The PMF and CDF of a discrete random variable

- The **probability mass function (PMF)** (or **probability distribution**) of a **discrete** random variable $X$ is

  <div class="box">
  $$f_X(x) = P(X = x); \quad  \sum_x f_X(x) = 1$$
  </div>
  
- The **cumulative distribution function (CDF)** of a discrete random variable $X$ is is

  <div class="box">
  $$F_X(x) = P(X \leq x)$$
  </div>

## The CDF of a continuous random variable

- The CDF of a **continuous** random variable $X$ is defined exactly as for discrete random variables:

  <div class="box">
  $$F_X(x) = P(X \leq x)$$
  </div>

## The probability density function

- The PMF is inappropriate for continuous random variables; instead, introduce the **probability density function (PDF)**

- The area under the PDF $f_X(X)$ b/w two points is equal to the probability that $X$ fall b/w those two points

  $$\int_a^b f_X(x) dx = F_X(b) - F_X(a)$$

- The PDF is the derivative of the CDF, and integrates to 1

  <div class="box">
  $$f_X(x) = \frac{d}{dx} F_X(x); \quad \int f_X(x) dx = 1$$
  </div>
  

## The Normal PDF and CDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
par(mfrow=c(2,1), mar = c(2,4.1,1,4.1))
# draw a plot of the N(0,1) PDF
x <- seq(-3.5,3.5,length=200)
pdf <- dnorm(x)
plot(x,pdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Density",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
# draw a plot of the N(0,1) CDF
x <- seq(-3.5,3.5,length=200)
cdf <- pnorm(x)
plot(x,cdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Probability",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

```


## The Normal PDF and CDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
par(mfrow=c(2,1), mar = c(2,4.1,1,4.1))
# draw a plot of the N(0,1) PDF

x <- seq(-3.5,3.5,length=200)
pdf <- dnorm(x)
plot(x,pdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Density",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

# The shaded area
xshade <- seq(-3.5,0,length=200)
pdfshade <- dnorm(xshade)
polygon(c(-3.5,xshade,0),c(0,pdfshade,0), col="lightgray", lwd = 0.1)
lines(x,pdf,
      lwd = 3,
      col="blue")

# draw a plot of the N(0,1) CDF
x <- seq(-3.5,3.5,length=200)
cdf <- pnorm(x)
plot(x,cdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Probability",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

segments(0, -1, 0, pnorm(0), col = "black", lty = 2, lwd = 3)
segments(-4, pnorm(0), 0, pnorm(0), col = "black", lty = 2, lwd = 3)
```


# Summary

## Summary

- A random variable assigns a number to a random outcome 

- The probabilities with which a discrete random variable takes on different values are given by the PMF

- The probabilities with which a continuous random variable takes on different values are given by the PDF

- The CDF gives the probability that a random variable takes a value less than or equal to a specific value
