---
pagetitle: Review of probability I
output: 
  revealjs::revealjs_presentation:
    incremental: false
    theme: solarized
    self_contained: false
    # reveal_plugins: ["menu","notes","chalkboard"]
    reveal_plugins: ["menu"]
    highlight: pygments
    center: true
    transition: none
    background_transition: none 
    reveal_options:
      # chalkboard:
      #   theme: whiteboard
      #   toggleNotesButton: true
      #   toggleChalkboardButton: true
      menu:
        numbers: true
      slideNumber: true
      previewLinks: false
    fig_caption: true
    css: mystyle.css
    
--- 

<section>

<h1>Random variables, distributions, and expectations</h1>

Based on Stock and Watson, ch. 2

<br>

<h2>[Jesper Bagger](mailto:jesper.bagger@rhul.ac.uk)</h2>

<h3>EC2203 | Royal Holloway | 2020/21</h3>



</section>

# Outline

- Random variables

- Distributions

- Expectations


# Random variables

## Randomness

- A random process is a definite outcome that cannot be predicted with certainty

  - Coin flip
  
  - Student test scores (for example, your EC2203 mark)
  
  - Waiting time in the College shop
  
  - The FTSE100 index
  
  - Salary in your first job
  
- We use **random variables** to analyze random processes
  
  
## Random variables

- Provides a **numerical** summary of a random outcome

- A random variable assigns a **real number** to every possible outcome of a random process

- A random variable takes **different values** depending on the outcomes of the random process under study


# Distributions

## Distributions

- The probability that, say, $X=x$ is the **proportion of times $X=x$ in many independent realizations** of $X$

- The **distribution** of a random variable specifies the probability of realizing each value of the random variable

- Distinguish b/w **discrete** and **continuous** random variables


## The PMF and CDF of a discrete random variable

- The **probability mass function (PMF)** (or **probability distribution**) of a **discrete** random variable $X$ is

  <div class="box">
  $$f_X(x) = P(X = x); \quad  \sum_x f_X(x) = 1$$
  </div>
  
- The **cumulative distribution function (CDF)** of a discrete random variable $X$ is is

  <div class="box">
  $$F_X(x) = P(X \leq x)$$
  </div>

## The CDF of a continuous random variable

- The CDF of a **continuous** random variable $X$ is defined exactly as for discrete random variables:

  <div class="box">
  $$F_X(x) = P(X \leq x)$$
  </div>

## The probability density function

- The PMF is inappropriate for continuous random variables; instead, introduce the **probability density function (PDF)**

- The area under the PDF $f_X(X)$ b/w two points is equal to the probability that $X$ fall b/w those two points

  $$\int_a^b f_X(x) dx = F_X(b) - F_X(a)$$

- The PDF is the derivative of the CDF, and integrates to 1

  <div class="box">
  $$f_X(x) = \frac{d}{dx} F_X(x); \quad \int f_X(x) dx = 1$$
  </div>
  

## The Normal PDF and CDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
par(mfrow=c(2,1), mar = c(2,4.1,1,4.1))
# draw a plot of the N(0,1) PDF
x <- seq(-3.5,3.5,length=200)
pdf <- dnorm(x)
plot(x,pdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Density",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
# draw a plot of the N(0,1) CDF
x <- seq(-3.5,3.5,length=200)
cdf <- pnorm(x)
plot(x,cdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Probability",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

```


## The Normal PDF and CDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
par(mfrow=c(2,1), mar = c(2,4.1,1,4.1))
# draw a plot of the N(0,1) PDF

x <- seq(-3.5,3.5,length=200)
pdf <- dnorm(x)
plot(x,pdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Density",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

# The shaded area
xshade <- seq(-3.5,0,length=200)
pdfshade <- dnorm(xshade)
polygon(c(-3.5,xshade,0),c(0,pdfshade,0), col="lightgray", lwd = 0.1)
lines(x,pdf,
      lwd = 3,
      col="blue")

# draw a plot of the N(0,1) CDF
x <- seq(-3.5,3.5,length=200)
cdf <- pnorm(x)
plot(x,cdf,
      type = "l",
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.5, 3.5),
      ylab = "Probability",
      xlab = " ",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)

segments(0, -1, 0, pnorm(0), col = "black", lty = 2, lwd = 3)
segments(-4, pnorm(0), 0, pnorm(0), col = "black", lty = 2, lwd = 3)
```


# Expectations

## The expected value of a random variable

- The **expected value** of $Y$, denoted $\textrm{E}(Y)$, is the average value of $Y$ over many repeated, independent occurrences 

  <div class="box">
  $$\textrm{E}(Y) = \left\{ 
  \begin{array}{ll}
  \sum_y y f_Y(y) & \text{if } Y \text{ discrete} \\[0.5ex]
  \int y f_Y(y) dy & \text{if } Y \text{ continuous}
  \end{array}
  \right.$$
  </div>

- $\textrm{E}(Y)$ is the **mean** of $Y$, and is sometimes denoted $\mu_Y$

## The mean $\mu_Y$ measures distribution location

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.,4 ),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 1, sd = 1),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("Mean = 0","Mean = 1")), 
       col = c("blue","red") , 
       lwd = 3,
      bty = "n",
       cex = 1.5)

```


## Variance of a random variable

- The **variance** of $Y$, denoted $\textrm{var}(Y)$, is the expected value of the squared deviation of $Y$ from its mean:

  <div class="box">
  $$\mathrm{var}(Y) = \textrm{E}[(Y-\mu_Y)^2]$$
  </div>

- The variance is often denoted $\sigma^2_Y$

## The variance $\sigma^2_Y$ measures distribution spread

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-6.5,6.5),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 0, sd = 2),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("Var = 1","Var = 4")), 
       col = c("blue","red") , 
       lwd = 3,
      bty = "n",
       cex = 1.5)

```


## The standard deviation of a random variable

- The **standard deviation** is the square root of the variance

  <div class="box">
  $$\sigma_Y = \sqrt{\sigma^2_Y}$$
  </div>

- The standard deviation is measured in units of $Y$


## A standardized random variable 

- **Standardization** transforms a random variable $Y$ into a random variable with mean 0 and variance 1

- The standardized random variable computed from $Y$ is

  <div class="box">
  $$\frac{Y - \mu_Y}{\sigma_Y}$$
  </div>

- Standardized random variables are **unit-free**

## The Normal distribution

- Suppose $Y$ has a **Normal distribution** with mean $\mu_Y$ and variance $\sigma_Y^2$. We write

  $$Y \sim \mathcal{N}(\mu_Y,\sigma_Y^2)$$
- If we standardize $Y$, we obtain random variable $Z$ with a **standard Normal distribution**; we write

  $$Z = \frac{Y-\mu_Y}{\sigma_Y} \sim \mathcal{N}(0,1)$$

- The standard Normal PDF is denoted $\phi(z)$; the standard Normal CDF is $\Phi(z) = \Pr(Z \leq z) = \int_{-\infty}^z \phi(x) dx$


## The standard Normal PDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-6.5,6.5),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 1, sd = 2),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("N(0,1)","N(1,4)")), 
       col = c("blue","red") , 
       lwd = 3,
       bty = "n",
       cex = 1.5)

```


# Summary

## Summary

- A random variable assigns a number to a random outcome 

- The probabilities with which a random variable takes on different values are summarized by the PMF/PDF (for discrete/continuous random variables) and the CDF

- The expected value of a random variable is the probability or density weighted average of the random variable

- The mean of a random variable measures the location of its distribution. 

- The standard deviation of a random variable measures the spread of its distribution.
